---
title: "A practical marcxmlr workflow"
author: "Lorenzo Isella"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A practical marcxmlr workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Overview

`marcxmlr` reads MARC 21 XML into a fixed 11-column canonical table that
preserves record structure, source order, repeated fields, repeated subfields,
and indicators. The same representation can be queried with ordinary R tools,
checked after editing, written back to MARCXML, or stored as Parquet for larger
collections.

This vignette follows one complete workflow:

1. read a MARCXML collection;
2. inspect repeated MARC structure;
3. query and derive analytical data with `dplyr`;
4. edit a repeated subfield;
5. diagnose and serialize the canonical representation; and
6. optionally move the same schema to a Parquet-backed workflow.

All examples use the same synthetic Orwell MARCXML record used in the README.
It is bundled with the package, so the vignette is offline and quick to build.

## Read MARCXML

```{r read}
library(marcxmlr)
library(dplyr)

source_xml <- system.file(
  "extdata",
  "orwell-marcxml.xml",
  package = "marcxmlr"
)

marc <- read_marcxml(source_xml)
marc
```

The result is an ordinary tibble with the canonical 11-column schema:

```{r schema}
names(marc)
```

The columns are `record_id`, `field_type`, `tag`, `subfield_code`, `value`,
`field_order`, `field_occurrence`, `ind1`, `ind2`, `subfield_order`, and
`subfield_occurrence`.

`record_id`, `field_order`, and `subfield_order` locate values in the represented
MARC structure. `field_occurrence` and `subfield_occurrence` are derived
analytical coordinates that make repetitions easier to address.

## Inspect repeated structure

The first record contains an `856` field with two `$y` subfields. Their source
order and separate occurrences remain explicit:

```{r repeated-subfields}
marc |>
  filter(record_id == 1L, tag == "856") |>
  select(
    subfield_code,
    value,
    subfield_order,
    subfield_occurrence
  )
```

Repeated fields are preserved in the same way. The first record contains two
`650` fields:

```{r repeated-fields}
marc |>
  filter(record_id == 1L, tag == "650", subfield_code == "a") |>
  select(record_id, field_order, field_occurrence, value)
```

This is why `marcxmlr` does not flatten MARCXML directly to one row per
bibliographic record: repeated field and subfield relationships remain explicit
until the analyst chooses how to simplify them.

## Query with dplyr

Because the canonical representation is a tibble, ordinary `dplyr` operations
can be used directly. For example, title fields can be selected by MARC tag:

```{r title-query}
marc |>
  filter(tag == "245") |>
  select(record_id, subfield_code, value)
```

A condition can also select a complete field instance rather than only the row
that matched. Rows with the same `record_id` and `field_order` belong to the
same field instance:

```{r complete-field}
marc |>
  group_by(record_id, field_order) |>
  filter(
    any(
      tag == "650" &
        subfield_code == "a" &
        value == "Totalitarianism"
    )
  ) |>
  ungroup() |>
  select(record_id, field_order, tag, subfield_code, value)
```

## Derive a simpler analytical table

The canonical representation preserves MARC structure first. A simpler table
can then be derived for a particular analysis.

```{r analytical-table}
titles <- marc |>
  filter(tag == "245", subfield_code == "a") |>
  transmute(record_id, title = value)

subjects <- marc |>
  filter(tag == "650") |>
  group_by(record_id, field_order) |>
  summarise(
    subject = paste(value, collapse = " ; "),
    .groups = "drop"
  ) |>
  group_by(record_id) |>
  summarise(
    subjects = paste(subject, collapse = " | "),
    .groups = "drop"
  )

left_join(titles, subjects, by = "record_id")
```

This simplification is deliberately one-way. Once field membership or
repetition has been collapsed, reconstructing the original MARC structure may
no longer be possible.

## Edit a repeated subfield

A specific repeated value can be addressed with the occurrence coordinates.
Here only the second `856$y` value in the first record is changed:

```{r edit}
edited <- marc |>
  mutate(
    value = if_else(
      record_id == 1L &
        tag == "856" &
        subfield_code == "y" &
        subfield_occurrence == 2L,
      "Backup access",
      value
    )
  )

edited |>
  filter(record_id == 1L, tag == "856") |>
  select(subfield_code, value, subfield_occurrence)
```

Only `value` has changed; the structural coordinates remain intact.

## Check before writing

After transformations that may affect MARC structure,
`diagnose_canonical()` can be used to inspect whether the representation is
still safe to serialize.

```{r diagnose}
diagnostics <- diagnose_canonical(edited)
diagnostics
```

A zero-row diagnostics tibble means that no issue was found. Errors identify
structural problems that prevent safe serialization. Warnings can identify
stale or gapped analytical occurrence coordinates that can be regenerated on
reread.

`diagnose_canonical()` checks the structural contract used by `marcxmlr`; it is
not a complete MARC 21 content validator and does not replace MARCXML schema
validation.

## Write MARCXML and read it back

`write_marcxml()` serializes coherent canonical data back to MARCXML. Structural
errors are rejected by the writer; with the default `check = TRUE`, warning-level
diagnostics are also emitted before writing.

```{r roundtrip}
out_xml <- tempfile(fileext = ".xml")
write_marcxml(edited, out_xml)

roundtrip <- read_marcxml(out_xml)
identical(edited, roundtrip)
```

A related workflow is to select complete records and export the result:

```{r subset-roundtrip}
selected <- marc |>
  filter(record_id == 1L)

selected_xml <- tempfile(fileext = ".xml")
write_marcxml(selected, selected_xml)

identical(selected, read_marcxml(selected_xml))
```

The round trip preserves the represented MARC structure and values. It does not
promise byte-for-byte reproduction of incidental XML serialization details such
as indentation or namespace-prefix spelling.

## Move the same schema to Parquet

For collections that are inconvenient to materialize as one R object,
`marcxml_to_parquet()` writes the same canonical schema as a directory of
Parquet files in bounded record batches. Arrow support is optional, so the
following examples run only when the optional packages needed for this workflow
are available.

```{r parquet, eval=requireNamespace("arrow", quietly = TRUE) && requireNamespace("XML", quietly = TRUE)}
parquet_dir <- tempfile("marcxmlr-parquet-")

marcxml_to_parquet(
  source_xml,
  output_dir = parquet_dir,
  workers = 1L,
  verbose = FALSE
)

catalogue <- arrow::open_dataset(parquet_dir)

catalogue |>
  summarise(rows = n()) |>
  collect()
```

The Parquet dataset can be queried lazily with Arrow and `dplyr` before a small
result is collected into R:

```{r parquet-query, eval=requireNamespace("arrow", quietly = TRUE) && requireNamespace("XML", quietly = TRUE)}
catalogue |>
  filter(tag == "650", subfield_code == "a") |>
  count(value, sort = TRUE) |>
  collect()
```

Opening the dataset is lazy; collecting the complete dataset would still bring
all rows into R memory.

## Choosing a workflow

For ordinary analysis, start with `read_marcxml()` and work with the returned
tibble. Use `diagnose_canonical()` after transformations that may affect
structure, and `write_marcxml()` when the result needs to return to MARCXML.

Use `marcxml_to_parquet()` when keeping the complete canonical table in memory
is undesirable or when a persistent Arrow/Parquet dataset is useful for repeated
analytical queries. The logical 11-column representation is the same in both
workflows.
